---
title: "Data Analysis with R"
author: "Fernando Hernandez"
date: "Monday, January 19, 2015"
output:
  html_document:
    highlight: espresso
    theme: journal
  pdf_document: default
geometry: margin=1in
fontsize: 10pt
---

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(gridExtra)
library(extrafont)
```

## Introduction

On May 30, 2014 MIT and Harvard released de-identified data from 13 HarvardX and MITx courses offered in 2012-13.[1]

The data contains records from 476,532 students covering 641,138 separate course registrations.[2]


## Initial Exploration

When exploring a new dataset, it's always a good idea to look at the structure with str(). 

This data set provides demographic and anonymized online activity data, including grades.

```{r load_data, warning=FALSE,message=FALSE}

setwd("data")
mooc_data <- tbl_df(read.csv("HMXPC13_DI_v2_5-14-14.csv", na.strings = c("NA", "")))

dim(mooc_data)

str(mooc_data)



```

The exact meanings of the variables can looked up in the "Person Course Documentation.pdf". First, let's tidy up the data a bit by separating course_id, which contains 3 different types of information, into 3 separate columns. We can also add Age using the user-provided birth year. [2]

We will then add the full title of the courses based on the Course Codes from the "Person Course Documentation.pdf".

```{r warning=FALSE,message=FALSE}
levels(mooc_data$course_id)

levels(mooc_data$gender) <- c("Female", "Male", "Other")

mooc_data <- mooc_data %>%
  separate(course_id, into = c("Institution", "Course_Code", "Semester"), sep="/", convert=TRUE) %>%
  rename(Country = final_cc_cname_DI,
         Level_of_Edu = LoE_DI,
         Registration_Date = start_time_DI,
         Last_Interaction = last_event_DI) %>%
    mutate(Age = 2014 - YoB,
         certified = factor(certified, labels=c("No", "Yes")),
         Level_of_Edu = factor(Level_of_Edu, levels = c("Less than Secondary", "Secondary", "Bachelor's", "Master's", "Doctorate")))

## Add in full course titles from codebook.[2]
course_names <- data.frame("Course_Code" = factor(c("CB22x", "CS50x", "ER22x", "PH207x", "PH278x", "14.73x",
                                             "2.01x", "3.091x", "6.002x", "6.00x", "7.00x", "8.02x", "8.MReV")),
                           "Full_Title" = c("The Ancient Greek Hero", "Intro to Computer Science I",
                                            "Justice", "Health in Numbers: Quant. Methods", 
                                            "Human Health & Global Envir. Change",
                                            "The Challenges of Global Poverty", "Elements of Structures",
                                            "Intro to Solid State Chemistry", "Circuits and Electronics",
                                            "Intro to C.S. and Programming", "Intro to Biology",
                                            "Electricity and Magnetism", "Mechanics Review"))

mooc_data <- left_join(mooc_data, course_names, by="Course_Code")

mooc_data <- mooc_data %>%
  mutate(letter_grade = cut(grade, breaks=c(0.00, 0.01, .49, .59, .69, .79, .89, 1.01), include.lowest = TRUE, ordered_result = FALSE))



# Add in School Colors for plots
# http://web.mit.edu/graphicidentity/colors.html
# http://www.hbs.edu/marketing/color.html

mooc_data <- mooc_data %>%
  mutate(School_Color = ifelse(Institution == "HarvardX", "#A41034", "#000000"))

school_colors <- mooc_data$School_Color
names(school_colors) <- mooc_data$Institution

```

## Univariate Exploration

This dataset contains data for 641,138 course registrations

476,532 unique students showing some students were registered for multiple courses in this dataset.

```{r warning=FALSE, message=FALSE}

table(mooc_data$registered)
str(levels(mooc_data$userid_DI))

```
About 400,262 (62.4%) of students accessed the Courseware tab which contains the videos, problem sets, and exams.
This leaves 240,876 students who never did anything in the course after registering.

```{r, message=FALSE, warning=FALSE}

prop.table(table(mooc_data$viewed))
table(mooc_data$viewed)
ggplot(mooc_data, aes(x=factor(viewed))) + geom_histogram() + scale_y_continuous(labels=comma)

```

Only about 39,686 (6.2%) of students accessed at least half of the chapters for a given course.

```{r, warning=FALSE,message=FALSE}

prop.table(table(mooc_data$explored))
table(mooc_data$explored)
ggplot(mooc_data, aes(x=factor(explored))) + geom_histogram() + scale_y_continuous(labels=comma)
```

Only 17,687 (2.8%) of registered students completed a given course and received a certificate (earned 50%-80% depending on course)

```{r, message=FALSE, warning=FALSE}
prop.table(table(mooc_data$certified))
table(mooc_data$certified)
ggplot(mooc_data, aes(x=factor(certified))) + geom_histogram() + scale_y_continuous(labels=comma)

```

The US and India have the largest amounts of registered students

```{r, warning=FALSE, message=FALSE}

ggplot(mooc_data, aes(x=factor(Country))) + geom_histogram() + scale_y_continuous(labels=comma) + coord_flip()

```

The most common education level of experience is a bachelors degree. 

```{r, message=FALSE, warning=FALSE}

prop.table(table(mooc_data$Level_of_Edu))
table(mooc_data$Level_of_Edu)
ggplot(mooc_data, aes(x=Level_of_Edu)) + geom_histogram() + scale_y_continuous(labels=comma)

```

Classes all start on different dates, this information doesn't seem too useful by itself

```{r, message=FALSE, warning=FALSE}

ggplot(mooc_data, aes(x=Registration_Date)) + geom_histogram() + scale_y_continuous(labels=comma) 

```


199,151 (31%) students had 0 interactions beyond just registering
171,714 (26.8%) students had 10 or less interactions with the website beyond registration

```{r, warning=FALSE, message=FALSE}

prop.table(table(mooc_data$nevents<=10, useNA="always"))
table(mooc_data$nevents<=10, useNA="always")

g1 <- ggplot(mooc_data, aes(x=nevents)) + geom_histogram() + scale_y_continuous(labels=comma) + scale_x_log10() + xlab("Log10 Scale")
g2 <- ggplot(mooc_data, aes(x=nevents)) + geom_histogram(binwidth=10) + coord_cartesian(xlim=c(0, 500)) + scale_y_continuous(labels=comma) + xlab("Binwidth = 10; limit 500")
grid.arrange(g1, g2)
```

457,530 (71.4%) did not play any videos at all within courses. 11.1% played videos 10 or less times, while 17.5% played videos more than 10 times.This is slightly startling because most of the instruction for courses are presented by the videos. One reason for the large proportion may be because some courses allow downloads of videos which are more effective for slower/intermittent internet connections.

```{r, message=FALSE, warning=FALSE}

prop.table(table(mooc_data$nplay_video<=10, useNA="always"))
table(mooc_data$nplay_video<=10, useNA="always")

g1 <- ggplot(mooc_data, aes(x=nplay_video)) + geom_histogram() + scale_y_continuous(labels=comma) + scale_x_log10(labels=comma) + xlab("Log10 Scale")
g2 <- ggplot(mooc_data, aes(x=nplay_video)) + geom_histogram(binwidth=10) + coord_cartesian(xlim=c(0, 500)) + scale_y_continuous(labels=comma) + xlab("Binwidth = 10; limit 500")
grid.arrange(g1, g2)

```

162,743 (25.4%) did not interact with the course any days at all.
422,441 (65.9%) had interaction on 10 or less unique days.
This isn't much considering courses typically run 2-3 months long.
55,954 (8.7%) interacted with the courses on more than 10 unique days.
This may be a good indicator for how well a student ultimately does.

```{r, message=FALSE, warning=FALSE}

prop.table(table(mooc_data$ndays_act<=10, useNA="always"))
table(mooc_data$ndays_act<=10, useNA="always")
ggplot(mooc_data, aes(x=ndays_act)) + geom_histogram() + scale_y_continuous(labels=comma)

```

40.4% of students didn't interact with any chapters of the course.
Number of chapters varies per course so this statistic may be difficult to interpret.
19% of students only interacted with 1 chapter.

```{r, message=FALSE, warning=FALSE}
prop.table(table(mooc_data$nchapters<=1, useNA="always"))
table(mooc_data$nchapters<=1, useNA="always")
ggplot(mooc_data, aes(x=nchapters)) + geom_histogram() + scale_y_continuous(labels=comma)

```

Very few students make any forum posts.
Only 7461 (1.2%) made any posts at all. 633,677 (98.8%) were either lurkers or did not visit the forums.

```{r,message=FALSE, warning=FALSE}

prop.table(table(mooc_data$nforum_posts==0, useNA="always"))
table(mooc_data$nforum_posts<=0, useNA="always")
ggplot(mooc_data, aes(x=nforum_posts)) + geom_histogram() + scale_y_continuous(labels=comma)

```

100,161 (15.6%) records showed some inconsistency between log tables of different variables.
This just shows that there is a bit of noise in the data, but overall it looks to be pretty consistent.

```{r, warning=FALSE, message=FALSE}

prop.table(table(mooc_data$incomplete_flag, useNA="always"))
table(mooc_data$incomplete_flag, useNA="always")

```

Mean age across the dataset is 28.75. Minimum of 1 and 592 people under 5 years old shows there may be some noise in the data. There are also 96,605 who did not enter a birth year.
Age shows a right skew with the midsection at 26-28, consistent with the largest group having a Bachelor's degree.


```{r, message=FALSE, warning=FALSE}

summary(mooc_data$Age)
table(mooc_data$Age<=5)
ggplot(mooc_data, aes(x=Age)) + geom_bar(binwidth=1, fill="orange", color="black") + scale_x_continuous(breaks=seq(0, 85, 5)) + scale_y_continuous(labels=comma) + ggtitle("Distribution of Ages for Registered Students")

```

Computer Science courses seem to be the most popular courses.

```{r, message=FALSE, warning=FALSE}
ggplot(mooc_data, aes(x=Full_Title)) + geom_histogram() + scale_y_continuous(labels=comma) + coord_flip()

```

MITx/HarvardX courses have roughly similar amounts of registered students in the dataset.

```{r,message=FALSE, warning=FALSE}

ggplot(mooc_data, aes(x=Institution)) + geom_histogram() + scale_y_continuous(labels=comma) + coord_flip()

```

##### Flagging Active Students as a feature.
Distributions may be very different for students who are at least somewhat active when compared to those who register but do little to nothing after.

```{r, warning=FALSE, message=FALSE}
ggplot(mooc_data, aes(x=grade)) + geom_freqpoly()
summary(mooc_data$grade)

# Limit to students who viewed the Courseware section at least once; viewed == 1 and
# had at least 1 interaction with the the course beyond registration; nevents > 0
mooc_data_active <- mooc_data %>%
  filter(viewed == 1 & nevents > 0)

ggplot(mooc_data_active, aes(x=grade)) + geom_freqpoly()

# Add the contraint that students had a final grade that was at least more than 0.0%; grade > 0.0
mooc_data_active <- mooc_data %>%
  filter(viewed == 1 & nevents > 0 & grade > 0.0)

ggplot(mooc_data_active, aes(x=grade)) + geom_freqpoly()

# Add the constraint that students explored at least 1/2 of the chapters of the course in the Courseware; explored == 1
mooc_data_active <- mooc_data %>%
  filter(viewed == 1 & nevents > 0 & grade > 0.0 & explored == 1)

ggplot(mooc_data_active, aes(x=grade)) + geom_freqpoly()
```

##### Based on this exploration, it might be a good idea to create a new variable, Active_User, based on course activity.

Active Users will be defined as registered users who:

1. Viewed the Courseware section at least once; viewed == 1
2. Had at least 1 interaction with the the course beyond registration; nevents > 0
3. Explored at least 1/2 of the chapters of the course in the Courseware; explored == 1
4. Had a final grade that was at least more than 0.0%; grade > 0.0

Note: #4 has some implications for HarvardX Intro to CS I, since that course is pass/fail (0.0%/100.0%) with a threshold of 50% to receive passing certificate.

```{r, warning=FALSE, message=FALSE}

# Add Active_User as a feature
mooc_data <- mooc_data %>%
  mutate(Active_User = ifelse( (viewed == 1 & nevents > 0 & explored == 1 & grade > 0.0), 1, 0),
         Active_User = factor(Active_User, labels=c("Not Active", "Active")))


```

There are a relatively small group who have no information for any of these features, but most students can be categorized as Active or No Active
There are 26,463 active students with 613,744 who were not active.
931 students had no information available in all 4 criteria.

```{r, message=FALSE, warning=FALSE}
table(mooc_data$Active_User, useNA="always")

```

About 4.1% of online students meet the criteria for being an active student

```{r, message=FALSE, warning=FALSE}
prop.table(table(mooc_data$Active_User))

ggplot(subset(mooc_data, !is.na(Active_User)), aes(x=Active_User)) + geom_histogram()

```

## Multivariate

```{r warning=FALSE,message=FALSE}

# Certified/Active
ggplot(subset(mooc_data, !is.na(Active_User)), aes(x=certified)) + geom_bar() + facet_wrap(~Active_User, scales = "free") + scale_y_continuous(labels=comma)

# Grades of Active Users

ggplot(subset(mooc_data, Active_User == "Active"), aes(x=grade)) + geom_freqpoly(aes(color=Institution), binwidth=0.01)
ggplot(subset(mooc_data, Active_User == "Active"), aes(x=grade)) + geom_freqpoly(aes(color=Full_Title), binwidth=0.01)
ggplot(subset(mooc_data, Active_User == "Active"), aes(x=grade)) + geom_freqpoly(aes(color=Country), binwidth=0.01)
ggplot(subset(mooc_data, Active_User == "Active"), aes(x=grade)) + geom_freqpoly(aes(color=gender), binwidth=0.01)
ggplot(subset(mooc_data, Active_User == "Active"), aes(x=grade)) + geom_freqpoly(aes(color=Level_of_Edu), binwidth=0.01)

```

A small of users below 10 years old looks suspect. 
There's an interesting split at 50%. 

```{r, message=FALSE, warning=FALSE}
ggplot(mooc_data, aes(x=Age, y=grade)) + geom_jitter(aes(color=Institution), alpha=1/10) + scale_y_continuous(labels=percent, breaks=seq(0, 1, 0.1)) + scale_x_continuous(breaks=seq(0, 80, 10)) + scale_color_manual(values = school_colors)

# Popularity
ggplot(mooc_data, aes(x=Full_Title)) + geom_histogram(aes(fill = Institution)) + theme(axis.text.x=element_text(angle=90, size=9)) + scale_fill_manual(values = school_colors) + coord_flip() + theme_minimal()

# Gender Enrollment
ggplot(subset(mooc_data, (!is.na(gender) & gender != "Other")), aes(x=Full_Title)) + geom_histogram(aes(fill=gender), position="dodge", color="black") + theme_minimal() + theme(axis.text.x=element_text(angle=90, size=9)) + coord_flip()

# Active User/Certification/Edu
ggplot(subset(mooc_data, !is.na(Active_User)), aes(x=Level_of_Edu)) + geom_bar(aes(fill=certified), position="dodge") + facet_wrap(~Active_User, scales = "free") + scale_y_continuous(labels=comma)

```

CS50 is a pass/fail 0%/100% course. Average scores are very low when all registered students are included. 
It may be better to look at active students to get a better feel for student performance since so many registered but didn't do anything in the course.


```{r,message=FALSE,warning=FALSE}
mooc_summary_any <- mooc_data %>%
  filter(!is.na(grade)) %>%
  group_by(Full_Title, Institution) %>%
  summarise(avg_grade_any = mean(grade),
            n_any = n())

mooc_summary_active <- mooc_data %>%
  filter(!is.na(grade),
         Active_User == "Active") %>%
  group_by(Full_Title, Institution) %>%
  summarise(avg_grade_active = mean(grade),
            n_active = n())

mooc_summary <- left_join(mooc_summary_any, mooc_summary_active) %>%
  mutate(active_perc = n_active/n_any)

ggplot(mooc_summary, aes(x=reorder(Full_Title, avg_grade_any), y=avg_grade_any)) + geom_bar(aes(fill=Institution), stat="identity") + theme(axis.text.x=element_text(angle=90, size=9)) + scale_fill_manual(values = school_colors) + coord_flip() + theme_minimal() + scale_y_continuous(labels=percent) + ggtitle("Among All Registered Users")

```

What about among Active Users? Since Intro to CS1 is pass/fail, anyone who is active does not have 0.0 as a grade, and must have full 100% causing the average to be 100%.

```{r, message=FALSE, warning=FALSE}
ggplot(mooc_summary, aes(x=reorder(Full_Title, avg_grade_active), y=avg_grade_active)) + geom_bar(aes(fill=Institution), stat="identity") + theme(axis.text.x=element_text(angle=90, size=9)) + scale_fill_manual(values = school_colors) + coord_flip() + theme_minimal() + scale_y_continuous(labels=percent) + ggtitle("Among Active Users")

```

An interesting thing to note here is that the most popular course for registrations, Intro to Science Science I, had by far the lowest percentage of active users.

```{r, warning=FALSE, message=FALSE}
ggplot(mooc_summary, aes(x=reorder(Full_Title, active_perc), y=active_perc)) + geom_bar(aes(fill=Institution), stat="identity") + theme(axis.text.x=element_text(angle=90, size=9)) + scale_fill_manual(values = school_colors) + coord_flip() + theme_minimal() + scale_y_continuous(labels=percent) + ggtitle("% of Active Users per Course")

```

Number of events and video plays don't seems to have a very high correlation with overall grades.

```{r, message=FALSE, warning=FALSE}
ggplot(mooc_data, aes(x=nevents, y=grade)) + geom_jitter(alpha=1/10)

ggplot(mooc_data, aes(x=nplay_video, y=grade)) + geom_jitter(alpha=1/10)

```

But number of unique days logged in and number of chapters viewed show some correlation.

```{r, message=FALSE, warning=FALSE}

ggplot(mooc_data, aes(x=ndays_act, y=grade)) + geom_jitter(alpha=1/10)

ggplot(mooc_data, aes(x=nchapters, y=grade)) + geom_jitter(alpha=1/10)

ggplot(mooc_data, aes(x=nforum_posts, y=grade)) + geom_jitter(alpha=1/10)

```

```{r, warning=FALSE, message=FALSE}

# Certified percentages
ggplot(mooc_data, aes(x=certified, y=..count../sum(..count..))) + geom_bar() + scale_y_continuous(labels=percent)

# Countries

country_grade <- mooc_data %>%
  filter(!is.na(grade)) %>%
  group_by(Country) %>%
  summarise(avg_grade = mean(grade))
# Average grade

ggplot(country_grade, aes(x=reorder(Country, avg_grade), y=avg_grade)) + geom_bar(stat="identity") + coord_flip()

# Education

# Among Active Users, there isn't much of a difference in final grades distributions
edu_grade <- mooc_data %>%
  filter(Active_User == "Active") %>%
  group_by(Level_of_Edu) %>%
  summarise(avg_grade = mean(grade))

ggplot(edu_grade, aes(x=Level_of_Edu, y=avg_grade)) + geom_bar(stat="identity")

# There are so many student with 0.0 grade, that anything above is an outlier; 3rd quantile is still 0.0!
summary(mooc_data$grade)
ggplot(mooc_data, aes(x=Level_of_Edu, y=grade, fill=Level_of_Edu)) + geom_boxplot()

# Just filtering out students with a grade of 0.0 shows the distributions better
ggplot(subset(mooc_data, grade > 0.0), aes(x=Level_of_Edu, y=grade, fill=Level_of_Edu)) + 
  geom_boxplot() +
  scale_y_continuous(breaks=seq(0, 1, .05)) +
  scale_fill_brewer(type="seq", palette=5) + 
  theme_minimal() +
  stat_summary(fun.y=median, colour="black", geom="text", show_guide = FALSE, vjust=-0.7, 
               aes( label=round(..y.., digits=2))) + 
  stat_summary(fun.y=mean, colour="black", geom="point", show_guide = FALSE, vjust=-0.7, shape=23, size=3)

table(mooc_data$letter_grade)

ggplot(subset(mooc_data, letter_grade != "[0,0.01]"), aes(x=letter_grade)) + geom_histogram() 

ggplot(subset(mooc_data, letter_grade != "[0,0.01]"), aes(x=letter_grade, y=Age)) + geom_boxplot()

ggplot(subset(mooc_data, letter_grade != "[0,0.01]"), aes(x=grade, color=Level_of_Edu)) + geom_density(position="dodge")


```

# Analysis

## Percentage Who Earned Completion Certificates

### How many students earned certificates?

Only 2.8% of registered students passed their respective courses.

```{r, warning=FALSE, message=FALSE}

cert <- mooc_data %>%
  filter(!is.na(certified)) %>%
  group_by(certified) %>%
  summarise(n=n()) %>%
  mutate(perc = n/sum(n))

ggplot(cert, aes(x=certified, y=perc, fill=certified)) + 
  geom_histogram(stat="identity", color="black") + 
  scale_y_continuous(labels=percent) +
  geom_text(aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")),
            color="black", size=6, vjust = -0.1) +
  scale_fill_manual(values=c("dodgerblue4", "goldenrod1")) + 
  ggtitle("Percentage of Registered Students\nThat Earned Certificates") + xlab("Earned Certification?") + ylab("Percentage Who Earned Certification") +
  theme(text=element_text(family="mono"))
  
```

Female students had a slightly higher course completion percentage, but overall, there wasn't a large difference in completion rate between the sexes.

```{r, message=FALSE, warning=FALSE}

mooc_gender <- mooc_data %>%
  filter(!is.na(gender),
         gender != "Other") %>%
  group_by(certified, gender) %>%
  summarise(n=n()) %>%
  group_by(gender) %>%
  mutate(percentage = n/sum(n))

ggplot(mooc_gender, aes(x=certified, y=percentage, fill=certified)) + geom_bar(stat="identity", color="black") +
  scale_y_continuous(labels=percent) + facet_wrap(~gender) + 
  geom_text(aes(family="mono", fontface="bold", y=percentage, label=paste(round(percentage, 3)*100, "%", sep="")),
            color="black", size=6, vjust = -0.1) +
  theme_minimal() + ggtitle("Percentage of Registered Students\nThat Earned Certificates by Gender") +
  scale_fill_manual(values=c("dodgerblue4", "goldenrod1")) +
  xlab("Earned Certification?") + ylab("Percentage Who Earned Certification") +
  theme(text=element_text(family="mono"))
```

When it comes to student location, Spain and Poland lead the pack for earning certificates of completion.

```{r, message=FALSE,warning=FALSE}
country_cert <- mooc_data %>%
  group_by(Country, certified) %>%
  summarise(n = n()) %>%
  group_by(Country) %>%
  mutate(perc = n/sum(n))

ggplot(subset(country_cert, certified=="Yes"), aes(x=reorder(Country, perc), y=perc)) +
  geom_bar(stat="identity", fill="#A31F34") +
  coord_flip() +
  scale_y_continuous(labels = percent) + 
  geom_text(aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")),
            color="white", size=3, hjust=1.5) + 
  theme_minimal() + ggtitle("Course Completion Percentage by Country") +
  theme(text=element_text(family="Calibri")) + ylab("Percentage Who Earned Certification") + xlab("Country")

```

## Course Enrollment

### What are the most popular courses?

Introductory courses for Computer Science and Programming are by far the most popular courses.

```{r, warning=FALSE, message=FALSE}
mooc_pop <- mooc_data %>%
  group_by(Full_Title, Institution) %>%
  summarise(n = n())

ggplot(mooc_pop, aes(x=reorder(Full_Title, n), y=n, fill=Institution)) +
  geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=90, size=9)) +
  scale_fill_manual(values = school_colors) + 
  coord_flip() + 
  theme_minimal() + 
  scale_y_continuous(labels=comma) + 
  ggtitle("Number of Registered Users per Course") + xlab("Course Name") + ylab("# of Regsitered Users") +
  theme(text=element_text(family="Calibri"))
```

Almost 3 times more males registered for courses than females.

But was there a difference in registered courses by gender?

```{r, warning=FALSE, message=FALSE}
ggplot(subset(mooc_data, (!is.na(gender) & gender != "Other")), aes(x=gender, fill=gender) ) +
  geom_bar(color="black") +
  theme_minimal() +
  scale_y_continuous(labels=comma) +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) + 
  ggtitle("Course Enrollment by Gender") + xlab("Gender") + ylab("Count") +
  theme(text=element_text(family="Calibri"))

```

The two most popular courses overall were also most popular for males. 

Courses closer to the Humanities nearly identical gender ratios while hard sciences were male-dominated in course registrations.

```{r, warning=FALSE, message=FALSE}
gender_enroll <- mooc_data %>%
  filter(!is.na(gender), 
         gender != "Other") %>%
  group_by(Full_Title, gender, Institution) %>%
  summarise(n = n()) 

ge1 <- ggplot(gender_enroll, aes(x=reorder(Full_Title, n), y=n, fill=gender) ) +
  geom_bar(stat="identity", position="dodge", color="black") +
  theme_minimal() +
  coord_flip() + 
  scale_y_continuous(labels=comma) +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) + 
  ggtitle("Course Enrollment by Gender") + xlab("Course Name") + ylab("# of Enrollments")

ge1

```

# Age

## What are the age distributions like for students?

Most registered students were between 20 and 28, with quite a few older students in a right-tailed distribution.

```{r, warning=FALSE, message=FALSE}

ggplot(mooc_data, aes(x=Age)) + 
  geom_bar(binwidth=1, fill="orange", color="black") + 
  scale_x_continuous(breaks=seq(0, 85, 5)) + 
  scale_y_continuous(labels=comma) +
  ggtitle("Distribution of Ages for Registered Students") + xlab("Age") + ylab("Count") + 
  theme(text=element_text(family="Calibri")) + 
  theme_minimal()

```

The U.S., Spain, Nigeria, and Japan had the oldest population of registered students while Morocco, Egypt, and India had the youngest populations of registered students.

```{r,message=FALSE,warning=FALSE}
country_age <- mooc_data %>%
  filter(!is.na(Age)) %>%
  group_by(Country) %>%
  summarise(avg_age = mean(Age)) 

ggplot(country_age, aes(x=reorder(Country, avg_age), y=avg_age)) +
  geom_bar(stat="identity", fill="#A31F34") +
  coord_flip(ylim=c(22, 34)) +
  geom_text(aes(family="mono", fontface="bold", y=avg_age, 
                label=round(avg_age, 2)), color="white", size=3, hjust=1.2) +
  ggtitle("Average Age for Registered Users by Country") + xlab("Country") + ylab("Average Age") + 
  theme(text=element_text(family="Calibri"))

```

# Education

Most students who register have a bachelor's degree. 

This remains true when filtering out only active users, although post-graduate degree percentages increase among active users.

```{r, warning=FALSE, message=FALSE}

edu <- mooc_data %>%
  filter(!is.na(Level_of_Edu)) %>%
  group_by(Level_of_Edu) %>%
  summarise(n=n()) %>%
  mutate(perc = n/sum(n))

e1 <- ggplot(edu, aes(x=Level_of_Edu, y=perc)) + 
  geom_bar(stat="identity", fill="#A31F34", color="black") +
  scale_y_continuous(labels=percent, limits=c(0, 0.50)) + 
  geom_text(aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")),
            color="black", size=5, vjust = -0.1)  +
  ggtitle("Percentage with each Education Level for All Registered Students") + ylab("Percentage") + xlab("Level of Education") +
  theme_minimal() + 
  theme(text=element_text(family="Calibri"))

edu2 <- mooc_data %>%
  filter(!is.na(Level_of_Edu),
         Active_User == "Active") %>%
  group_by(Level_of_Edu) %>%
  summarise(n=n()) %>%
  mutate(perc = n/sum(n))

e2 <- ggplot(edu2, aes(x=Level_of_Edu, y=perc)) +
  geom_bar(stat="identity", fill="#A31F34", color="black") +
  scale_y_continuous(labels=percent, limits=c(0, 0.50)) + 
  geom_text(aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")),
            color="black", size=5, vjust = -0.1) + 
  ggtitle("Percentage with each Education Level for Active Users") + ylab("Percentage") + xlab("Level of Education") +
  theme_minimal() + 
  theme(text=element_text(family="Calibri"))

grid.arrange(e1, e2)
```

Within each gender, males have noticeably higher percentages of registered students in Secondary school and while female have higher percentages of students with master's degrees. This difference becomes more pronounced among active users.

```{r, message=FALSE, warning=FALSE, fig.height=8, fig.width=12}

mooc_edu1 <- mooc_data %>%
  filter(!is.na(Level_of_Edu),
         !is.na(gender),
         gender != "Other") %>%
  group_by(gender, Level_of_Edu) %>%
  summarize(n=n()) %>%
  mutate(perc = n/sum(n))

e3 <- ggplot(mooc_edu1, aes(x=Level_of_Edu, y=perc, fill=gender)) + geom_bar(stat="identity", position="dodge", color="black") +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) +  
  geom_text(data=subset(mooc_edu1, gender=="Female"), 
            aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")), 
            color="#CC79A7", size=4, vjust = -0.2, hjust=1.3) +
  geom_text(data=subset(mooc_edu1, gender=="Male"), 
            aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")), 
            color="#56B4E9", size=4, vjust = -0.2, hjust=-0.3) +
  scale_y_continuous(labels=percent, breaks=seq(0,0.45,0.05), limits=c(0, 0.45)) + 
  theme_minimal() + 
  theme(text=element_text(family="Calibri")) + 
  ggtitle("Percentage with each Education Level within each Gender for All Registered Students") +
  ylab("Percentage within Gender") + xlab("Level of Education")

mooc_edu2 <- mooc_data %>%
  filter(!is.na(Level_of_Edu),
         !is.na(gender),
         Active_User == "Active",
         gender != "Other") %>%
  group_by(gender, Level_of_Edu) %>%
  summarize(n=n()) %>%
  mutate(perc = n/sum(n))

e4 <- ggplot(mooc_edu2, aes(x=Level_of_Edu, y=perc, fill=gender)) + geom_bar(stat="identity", position="dodge", color="black") +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) +  
  geom_text(data=subset(mooc_edu2, gender=="Female"), 
            aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")), 
            color="#CC79A7", size=4, vjust = -0.2, hjust=1.3) +
  geom_text(data=subset(mooc_edu2, gender=="Male"), 
            aes(family="mono", fontface="bold", y=perc, label=paste(round(perc, 3)*100, "%", sep="")), 
            color="#56B4E9", size=4, vjust = -0.2, hjust=-0.3) +
  scale_y_continuous(labels=percent, breaks=seq(0,0.45,0.05), limits=c(0, 0.45)) + 
  theme_minimal() + 
  theme(text=element_text(family="Calibri")) + 
  ggtitle("Percentage with each Education Level within each Gender for Active Students") +
  ylab("Percentage within Gender") + xlab("Level of Education")

grid.arrange(e3, e4)

```

# Grades

And finally, we investigate how well do students do in the courses.

First, an age-old question: Who gets better grades - men or women?

Among active users, it's very close to an outright tie with only 1.8% difference in overall average grades.

```{r, message=FALSE, warning=FALSE}
mooc_gender_active_grade <- mooc_data %>%
  filter(!is.na(gender),
         gender != "Other",
         Active_User == "Active") %>%
  group_by(gender) %>%
  summarise(avg_grade = mean(grade, na.rm=TRUE),
            n=n())

ggplot(mooc_gender_active_grade, aes(x=gender)) + geom_bar(aes(fill=gender, y=avg_grade), stat="identity") +
  coord_flip() + 
  scale_y_continuous(labels=percent) +
  geom_text(aes(family="mono", fontface="bold", y=avg_grade, label=paste(round(avg_grade, 3)*100, "%", sep="")),
            color="white", size=10, hjust=1) + 
  ggtitle("Average Grade by Gender for Active Users") + ylab("Average Grade") + xlab("Gender") +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) + 
  theme_minimal() + 
  theme(text=element_text(family="Calibri"))
```

But with different genders being drawn to different courses, how did grades differ between the different classes?

Women did slightly better in many of the classes that were dominated by male registrants. It also seems that the larger numbers of males passing Intro to CS I (100% pass/fail course) may have helped boost overall male grade % a bit.

We can also see that Electricity and Magnetism was the hardest course for active students.

Note: Again, Intro to CS I was pass/fail so the average grade for active users here is 100%.

```{r, message=FALSE, warning=FALSE}
mooc_gender_active <- mooc_data %>%
  filter(!is.na(gender),
         gender != "Other",
         Active_User == "Active") %>%
  group_by(Full_Title, gender, Institution) %>%
  summarise(avg_grade = mean(grade, na.rm=TRUE),
            n=n())

ge2 <- ggplot(mooc_gender_active, aes(x=reorder(Full_Title, avg_grade), y=avg_grade, fill=gender) ) +
  geom_bar(stat="identity", position="dodge", color="black") +
  theme_minimal() +
  coord_flip() + 
  scale_y_continuous(labels=percent, breaks=seq(0, 1, 0.1)) +
  scale_fill_manual(values=c("#CC79A7", "#56B4E9")) + 
  ggtitle("Average Grades per Course by Gender for Active Users") + xlab("Course Name") + ylab("Average Grade in Course") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ge2
```

Averages grades by age for active users stayed fairly consistent until around 50 years of age where average grades become erratic. This is probably due in part to the right-skew of the data causing there to be less data available for ages 50+.

We can see the min/max in the lighter bands of color, while the IQR is represented by the darker bands.

```{r, message=FALSE, warning=FALSE}
age_grades <- mooc_data %>%
  filter(!is.na(grade),
         Active_User == "Active",
         Age > 5) %>%
  group_by(Age) %>%
  summarise(avg_grade = mean(grade),
            median_grade = median(grade),
            sd = sd(grade),
            lower = min(grade),
            upper = max(grade),
            quant.25 = quantile(grade, probs=0.25, na.rm=TRUE),
            quant.75 = quantile(grade, probs=0.75, na.rm=TRUE),
            n = n())

p <- ggplot(age_grades, aes(x=Age, y=avg_grade)) + 
  theme(plot.background = element_blank(),
                panel.grid.minor = element_blank(),
                panel.grid.major = element_blank(),
                panel.border = element_blank(),
                panel.background = element_blank(),
                axis.ticks = element_blank(),
                axis.title = element_blank()) +
  geom_linerange(age_grades, mapping=aes(x=Age, ymin=lower, ymax=upper), colour = "wheat2", alpha=1, size=2.5) + 
  geom_linerange(age_grades, mapping=aes(x=Age, ymin=quant.25, ymax=quant.75), colour = "wheat4", size=2.5) +
  geom_line(age_grades, mapping=aes(x=Age, y=avg_grade, group=1)) +
  geom_vline(xintercept = 12, colour = "wheat4", linetype=1, size=1, hjust=0) + 
  geom_hline(yintercept=seq(0, 1, 0.1), color="white", linetype=1) 

dottedAges <- seq(15, 75, 10) 

p <- p + geom_vline(xintercept = dottedAges, color="wheat4", linetype=3, size=0.5)

p <- p + coord_cartesian(ylim = c(0,1.02)) +
  scale_y_continuous(labels=percent) +
  scale_x_continuous(breaks=seq(12, 82, 5), expand=c(0,0) ) +
  ggtitle("Grades") +
  theme(plot.title=element_text(face="bold",hjust=0,vjust=.8,color="#3C3C3C",size=20)) + 
  annotate("text", x=15, y=1, label="by Age", size=4, fontface="bold") + 
  annotate("text", x=32, y=.21,   label="Data represents average grades at each age from ages", size=3.5, color="gray30") + 
    annotate("text", x=32, y=.18, label="12 to 79. Lighter outer bands show the min/max grade ", size=3.5, color="gray30") +
    annotate("text", x=32, y=.15, label="for each age, and darker inner bands show the IQR.  ", size=3.5, color="gray30") +
  geom_point()

p
```

Averages grades for the different levels of education for active users showed large bump for those who have a post-graduate degree.


```{r, message=FALSE, warning=FALSE}
ggplot(subset(mooc_data, Active_User == "Active" & !is.na(Level_of_Edu)), aes(x=Level_of_Edu, y=grade, fill=Level_of_Edu)) +
  geom_boxplot() +
  scale_y_continuous(breaks=seq(0, 1, .05), labels=percent) +
  scale_fill_brewer(type="seq", palette=5, name="Education Level", guide=FALSE) + 
  theme_minimal() + 
  stat_summary(fun.y=median, colour="black", geom="text", show_guide = FALSE, vjust=-0.7,
               aes( label=paste(round(..y.., 4)*100, "%", sep=""))) + 
  stat_summary(fun.y=mean, colour="black", geom="point", show_guide = FALSE, vjust=-0.7, shape=23, size=3) + 
  ggtitle("Boxplots of Grades by Level of Education for Active Students") + 
  ylab("Grade") + xlab("Level of Education") +
  theme(text=element_text(family="Calibri"))

```

If we look at just those who earned a certificate, this progression in average grades by education level continues.

```{r, message=FALSE,warning=FALSE}
ggplot(subset(mooc_data, certified == "Yes" & !is.na(Level_of_Edu)), aes(x=Level_of_Edu, y=grade, fill=Level_of_Edu)) + 
  geom_boxplot() + 
  scale_y_continuous(breaks=seq(0, 1, .05), labels=percent) + 
  scale_fill_brewer(type="seq", palette=5, name="Education Level", guide=FALSE) +
  theme_minimal() + 
  stat_summary(fun.y=median, colour="black", geom="text", show_guide = FALSE, vjust=-0.7,
               aes( label=paste(round(..y.., 4)*100, "%", sep=""))) +
  stat_summary(fun.y=mean, colour="black", geom="point", show_guide = FALSE, vjust=-0.7, shape=23, size=3) +
  ggtitle("Boxplots of Grades by Level of Education for Those Who Earned a Certificate") + 
  ylab("Grade") + xlab("Level of Education") + 
  theme(text=element_text(family="Calibri"))
```

And finally, grades for active users by country show Germany, Russian Federation, Portugal and Spain at the head of the class.

```{r, message=FALSE, warning=FALSE}
country_grade <- mooc_data %>%
  filter(!is.na(grade),
         Active_User == "Active") %>%
  group_by(Country) %>%
  summarise(avg_grade = mean(grade))

ggplot(country_grade, aes(x=reorder(Country, avg_grade), y=avg_grade)) + 
  geom_bar(stat="identity", fill="#A31F34") +
  coord_flip() +
  scale_y_continuous(labels = percent) + 
  geom_text(aes(family="mono", fontface="bold", y=avg_grade, 
                label=paste(round(avg_grade, 3)*100, "%", sep="")), color="white", size=3, hjust=1.5) + 
  theme_minimal() + 
  ggtitle("Average Grades by Country for Active Users") + ylab("Average Grade") + xlab("Country")

```


# Final Plots

```{r, warning=FALSE, message=FALSE}


```


```{r, warning=FALSE, message=FALSE}

grid.arrange(ge1, ge2, ncol=1)

```


```{r, warning=FALSE, message=FALSE}

```


[1] http://newsoffice.mit.edu/2014/mit-and-harvard-release-de-identified-learning-data-open-online-courses

[2] http://thedata.harvard.edu/dvn/dv/mxhx
